---
title: "Visualizing An Endpoint Metric With Variation"
author: P. Lombardo and G. Kwan
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(dplyr)
library(ggplot2)
knitr::opts_chunk$set(echo = FALSE)
experiment <- read.csv('data/experiment_data.csv', header = T)
```


## Experimental Set Up
Review experiment and data

### Exploring data set
have them explore with `head(experiment)`

Overview the goals of the lesson, then

* link to tidy-ing videos and lessons, and 
* ggplot basics for incremental building of plots.

###  Preparing endpoint data, saving as `df`
testing stuff

## Planning our plot
An example picture, but "hand-drawn"


Which data?
What are the axes?
What "geometries" appear?

## Incrementally building our plot

#### Start the `ggplot()` with `data =` and `aes()`thetics

#### Add `geom_*`-etries to visualize the data

#### `scale_*` your `aes()`thetics 

## Finalizing the plot

#### Some other adjustments

#### Add a title and axis labels with `labs()`

#### Optional: overlay the data!

## Review Exercises

## Optional: Modified versions of this plot
Show a few variations, highlighting changes made in the code to creat the modifications:



***
## Example exercises

### Exercise 

*Here's a simple exercise with an empty code chunk provided for entering the answer.*

Write the R code required to add two plus two:

```{r two-plus-two, exercise=TRUE}

```

### Exercise with Code

*Here's an exercise with some prepopulated code as well as `exercise.lines = 5` to provide a bit more initial room to work.*

Now write a function that adds any two numbers and then call it:

```{r add-function, exercise=TRUE, exercise.lines = 5}
add <- function() {
  
}
```

## Topic 2

### Exercise with Hint

*Here's an exercise where the chunk is pre-evaluated via the `exercise.eval` option (so the user can see the default output we'd like them to customize). We also add a "hint" to the correct solution via the chunk immediate below labeled `print-limit-hint`.*

Modify the following code to limit the number of rows printed to 5:

```{r print-limit, exercise=TRUE, exercise.eval=TRUE}
mtcars
```

```{r print-limit-hint}
head(mtcars)
```


### Shiny Example
```{r, echo=FALSE}
sliderInput("bins", "Number of bins:", min = 1, max = 50, value = 30)
plotOutput("distPlot")
```

```{r, context="server"}
output$distPlot <- renderPlot({
  x <- faithful[, 2]  # Old Faithful Geyser data
  bins <- seq(min(x), max(x), length.out = input$bins + 1)
  hist(x, breaks = bins, col = 'darkgray', border = 'white')
})
```

### Quiz

*You can include any number of single or multiple choice questions as a quiz. Use the `question` function to define a question and the `quiz` function for grouping multiple questions together.*

Some questions to verify that you understand the purposes of various base and recommended R packages:

```{r quiz}
quiz(
  question("Which package contains functions for installing other R packages?",
    answer("base"),
    answer("tools"),
    answer("utils", correct = TRUE),
    answer("codetools")
  ),
  question("Which of the R packages listed below are used to create plots?",
    answer("lattice", correct = TRUE),
    answer("tools"),
    answer("stats"),
    answer("grid", correct = TRUE)
  )
)
```

